---
title: "Challenging unifactorial status in CFA measurement models"
short_title: "Correlation testing after factor exclusion"
short_author: "A. Achim"
authors: 
- name: "André Achim"
  affiliation: "Université du Québec à Montréal"
  city: "Montréal"
  country: "Canada"
  email: "achim.andre\\@uqam.ca"
  address: "Département de Psychologie, C.P. 8888, succursale Centre-ville, Montréal (Québec) H3C 3P8"
  ORCID: "0000-0003-2159-6010"
abstract: |
  Confirmatory factor analysis (CFA) adjusts the coefficients of a hypothesised factor model to optimally fit the data. The initial measurement model often assigns most variables to only one of the common factors. With sufficient power, missing cross-factor links should result in a significant $\chi^2$ fit index, accompanied by suggestions for improving the fit. *Residual correlation* is introduced to detect correlations outside a factor direction, either ahead of CFA optimization or to support suggested additional links beyond mere statistical acceptability. The factor is specified from two variables initially documented exclusive to a common factor. A remarkable property of such factor exclusive variables allows extracting from the correlation matrix the projections of all remaining variables on that factor. These allow excluding the factor contributions from all correlations, leaving residual correlations. Variables exclusively informed by the target factor behave as noise with no significant residual correlation. Any variable assigned to the target factor but having a significant residual correlations with indicators of other factors cannot be assumed unifactorial. After simulations document the statistical behaviour of residual correlation testing, a real-data textbook example illustrates that not one, but two of the Verbal IQ indicators are further informed by the Performance IQ factor.
keywords: Conformatory factor analysis; Factor exclusion; Measurement model; Residual correlation; Signal cancellation
# bibliography: bibliography.bib
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage[T1]{fontenc}
  - \lstset{basicstyle=\small\ttfamily, breaklines=false}
  - \usepackage{float}
  - \usepackage{setspace}
  - \doublespacing
  - \geometry{a4paper}
  - \renewcommand{\maketitle}{}

output:
  rticles::rss_article:
    latex_engine: pdflatex
    includes:
      in_header: fix_abstract.tex
---

# Introduction

The common factor model for $n_f$ factors expressed in $n_v$ noisy variables
acknowledges that $n_f+n_v$ sources of information are expressed in only $n_v$
directly accessible dimensions. Confirmatory factor analysis (CFA) aims at
isolating the $n_f$ factors from the $n_v$ sources of unique variance based on
a postulated factor model. These models frequently assign variables to only one
of the postulated factors. Even with the correct number of factors, the model
may err by missing cross-factor links, resulting in rejecting the compatibility
of the data with the factor model, expressed as a significant $\chi^2$ fit
index. Most CFA software will suggest model modifications to improve the fit.
Parsimony dictates adding no more links than necessary to make the model
statistically acceptable. A real data textbook example will illustrate that the
method introduced here may provide information to justify adding links beyond
bringing the model to non-significance.

Alternately, rather than adding links post initial CFA optimisation, it is
possible to pretest whether cross-factor links are missing. Both contexts
involve excluding a factor direction from all variables and testing whether
each variable thought exclusively informed by that factor indeed behaves as
noise outside this direction. The definition of noise orthogonality requires
that the factor direction be outside the subspace of the *nv* sources of unique
variance, which partial correlation cannot achieve by linearly combining noisy
variables. Such combinations are correct for prediction but not for inferential
purposes unless an independent variable expresses the factor without
measurement error. The present factor exclusion approach rather relies on the
common signal of a pair of noisy variables each exclusive to the same factor.

If two variables are exclusive to the same factor, their respective signal parts
have the direction of this factor while the variables themselves define two different 
directions due to the contribution of their unique sources of
variance. The shared direction of their
signal implies that one variable may be weighted to make its
signal vector identical to that of the other and subtracted from the latter
forming a contrast with no signal, consisting only in a weighted sum of their
respective noise components. The suitable signal cancellation weight is the
value that makes the contrast uncorrelated with any other variable. Successful
signal cancellation between two variables allows using them to designate 
their common factor to test for shared information outside its direction. 
The statistics is designated *residual correlation*.

# Residual Correlation Principles

Besides the usual power considerations, the residual correlation approach only
requires that the target factor can be designated by a pair of variables, say
*A* and *B*, that are exclusive to it except for their respective unique
sources of variance. Their exclusiveness to the same common factor implies that
a constant weight *w* exists such that the contrast *A-wB* cancels their
signal. Such contrast then exclusively consists of their combined unique
sources of variance, void of common factor information. Computationally, the
suitable weight *w* minimises the multiple regression of the *A-wB* contrast on
all remaining variables. The resulting $R^2$ must be clearly not significant to
validate that the contrast consists of noise only, i.e. that the contrast
successfully cancelled the signal from the only common factor informing *A* and
*B*. For other pairs of variables, including *A* with *X* and *B* with *X*, a
still significant minimised $R^2$ indicates failure of signal cancellation,
meaning that the variables contrasted express at least two factors, their
respective signal parts not being aligned on a single common factor.

Respectively labelling *a* and *b* the factor loadings of *A* and *B*, the
optimal weight *w* must correspond to $a/b$, where the denominator re-scales
*B*'s signal to unit length, which the numerator then matches to *A*'s signal
length for signal cancellation. Following successful signal cancellation, the
two loadings are easily estimated since *w* is the ratio of *a* and *b* while
$\rc{AB}$ is their product. Solving these two theoretical equalities yields $a^2
= \rc{AB}w$ and $b^2 = \rc{AB}/w$, leaving only the signs of *a* and *b* to be
attributed.
```{r}

```

For the present purpose, it suffices to use $\rc{AB}$ as the product of *a* and
*b*. The projection of each remaining variable on their common factor is
extracted from its correlations with each of *A* and *B*. Labelling $x_1$ the
projection of a generic variable *X* on that factor, the expected correlations
$\rc{AX}$ and $\rc{BX}$ must be $ax_1$ and $bx_1$ respectively since *A* and *B*
load zero on all remaining common factors. It is then readily seen that $x_1^2$
is $\rc{AX} \rc{BX}/\rc{AB}$, which may also be written as $(ax_1)(bx_1)/(ab)$
that decomposes as $(ab)x_1^2/(ab)$. The sign of $x_1$ is then set consistent
with the observed correlations involved.

The residual correlation of any pair of variables, i.e. the contribution to
correlation outside the target factor direction, is then obtained by
subtracting from their full correlation the product of their loadings on the
target factor direction. Indeed, all correlations being sums of coordinate
products over any convenient orthogonal basis, the subtraction partitions such
sum of products into the product of the estimated loadings on the designated
factor direction and a residual sum of products.

The resulting residual correlation is tested as the difference between the
original correlation and its single factor estimate, each subjected to Fisher's
$z_r$ transformation (atanh). The difference is $D=z(\rc{XY} )-z({xy})$, where
*x* and *y* are the projections of variables *X* and *Y* on the excluded
factor, with their product *xy* calculated as $\sqrt{\rc{AX}\rc{BX}\rc{AY}\rc{BY}}
/\rc{AB}$. The contrast *D*, expected null when the designated factor is the
only common contributor to *X* and *Y*, is thus a function of six correlated
correlations. The asymptotic covariance matrix of the six sample correlations
is obtained from the Pearson-Filon formula (Steiger, 1980). The gradient of *D*
with respect to these six correlations — which incorporates the derivatives of
the atanh transformation — is then used to propagate their joint sampling
variability into the variance of *D* via the delta method: $\text{Var}(D) =
\mathbf{g}^\top \Sigma \mathbf{g}$. The test statistics divides *D* by its
standard error, referring the result to the normal distribution to reflect that
$Var(D)$ is derived asymptotically.

# Simulation Study

## Methods

The main aim of the simulation study is to document compliance with the nominal
false positive error rate. It is secondarily to document eventual bias by
comparing the estimated residual correlations to those expected from the
population parameters. More specific to the new method, the third aim is to
document how the correlation between the factor indicator variables (FIVs)
interacts with the residual correlation effect size. To document these three aspects, the study design varies 
sample size and FIV loadings over a range of expected
residual correlations.

Since residual correlation concerns relationships among four variables, namely
the tested pair of dependent variables and the two FIVs, a single factor
structure may assess a variety of conditions. Besides incorporating a range of
effect sizes, the design implements three types of expected null residual
correlations. First, one of the two dependent variables may consist exclusively
of noise outside the excluded factor. Next, pairs of bifactorial dependent
variables may be informed by the target factor and by a different second
factor. Finally, a pair of dependent variables may have loadings forming a null
sum of products outside the excluded direction. 

A three-factor structure was designed to embed the three types of null
expected residual correlations besides forming a range of effect sizes. Within
it, three variables load exclusively on the factor to be excluded. The first
two of these, serving as FIV, are given loadings that vary across runs from
(.4, .5) to (.8, .9) by steps of (.1, .1). The third variable that loads
exclusively on the target factor has expected null residual correlations with
all remaining variables. The remaining variables form pairs that either exemplify the
other two types of null expected residual correlations or have expected
residual correlations in the range .05 to .34.

Three factors were originally defined to correlate .4 between $F_1$ and $F_2$,
.5 between $F_1$ and $F_3$, and .6 between $F_2$ and $F_3$. Since factor
exclusion is the removal of projections on the direction of a target factor, it
was convenient the express the factor space on orthogonal axes through the
Cholesky decomposition of the factor correlation matrix. This keeps the target
factor direction as the first axis and makes the sums of pairwise
cross-products loadings over the last two axes the expected residual
correlations. The test factor structure, finalised by rounding all loadings to
two decimal places, is presented in Table~\ref{tab:structure}. All variables
have a communality below .9. The 11 dependent variables $v_3$ to $v_{14}$
provide 66 residual correlation estimates, 20 of which are expected null.

No correction for the number of
tests is applied considering that this would make the statistical results of
any combination of parameters contingent on the current context rather that to
document specific combinations that could occur with actual data. Nonetheless,
the default .05 $\alpha$ level is replaced by a 99% confidence interval for
each estimation of false positive rates.

With each set of FIV loadings, sample size was set at 125, 500 or 2000, where
this number also served to initialise the pseudo-random number generator. This
ensured identical data for all variables other than the FIVs at any given
sample size. For each of these 15 combinations of parameters, a minimum of 1000
data sets were used, replacing those in which the FIV failed the imposed
requirement that they correlate significantly at the .05 level, which occurred
mostly for $N=125$ with the expected correlation of .20 and, occasionally, with
the expected correlation of .30.

Out of the 1000 data sets subjected to residual correlation analysis, a few
could involve an assessed loading above unity (Heywood case), again occurring
essentially with the lowest FIV loadings and the smallest sample size. The
warning to this effect was disabled for the simulations. This would normally
lead a user to ignore the results or to exclude the variable that received a
loading above unity. The information being only available after residual
correlation analysis, these rare improper data sets were not replaced but
ignored in calculating the proportion of significant residual correlations.

## Results

### False Positive Rates
The upper limit of the 99% confidence interval for 1000 tests performed at the
.05 level is 6.8%. The method was seen to respect the nominal false positive
rate except in the hardest conditions. Across the 20 pairs of variables with
expected null residual correlations, the false positives ranged from 3.2% to
6.8% (with one exception at 7.1% that became 5.1% upon an independent
replication) in all conditions but the first one with the smallest sample size
and the smallest FIV loadings.

With FIV loadings of (.4, .5) along with $N=125$, nine excessive rates were
seen, although all below 10%. These occurred with all three types producing null residual correlations. Of the 11 type-1 pairs involving $v_3$
(exclusively informed by the excluded factor), those formed with $v_4$, $v_9$,
$v_{10}$, and $v_{14}$ respectively showed 7.6%, 9.0%, 7.9%, and 7.6% false
positives. Three of the six type-2 pairs informed by different factors outside
the excluded one gave excess false positives. These are the crossings of $v_4$
with $v_7$ or $v_8$ and the crossing of $v_6$ with $v_7$ that respectively gave
8.6%, 7.7% and 7.9% false positives. Finally, for type-3, $v_{13}$ gave 9.2% false
positives when crossed with either $v_9$ or $v_{14}$ (but only 4.0% crossed
with $v_{12}$).

An independent replication with these smallest FIV correlation and sample size
suggested much variability in exceeding or not 7% false positives. The
$v_6-v_7$ pair shifted from 7.9% to 6.4% while three other pairs shifted upward
from 5.4% to 8.5%, from 6.2% to 8.2%, and from 6.8% to 7.6%.

### Residual Correlations Values
The black line in Figure~\ref{fig_1} represents the 46 population residual
correlations placed in increasing order. Equal expected values are ordered in
increasing order of overall observed values. The white lines are the means
estimates over the 1000 or so data sets (i.e. excluding the occasional ones
with a Heywood case) of each condition. The grey ribbon represents one standard
deviation from the mean. Given the large number of independent estimates, the standard error
of the mean is quite small. The mean estimates whose 99.9% confidence interval
does not include the expected value are marked by a pale grey vertical bar with
its centre on the observed mean. All computations were carried on Fisher
*atanh* transformed values but are painted back-transformed to the original
scale. Two negative residual correlations, at -.12 and -.18, are sign-reversed
for clarity.

The expected value is always seen within one standard deviation of the near
1000 datasets of each condition. The largest sample size and largest loadings
on the two variables designating the excluded factor indicate excellent
asymptotic recovery of the population residual correlations.

### Effect of Indicator Variable Loadings on Detection Power.
Although there is no
doubt that detection power depends on effect and sample sizes, the contribution
of the FIV signal level remains to document. Due to the resetting of the
pseudo-random number generator to sample size, the simulated data were strictly
paired within sample size. While the FIV loading levels could be compared
pairwise by McNemar's method or collectively by Cochran's Q test separately for
each sample size and each of the 46 pairs of dependent variables or on their
totals, their GLM extension, namely the conditional logistic regression
(Therneau & Grambsch, 2000; 'clogit' function of the survival R package,
Therneau, 2026) allows testing the effect of FIV loading levels across the
three sample sizes and the 46 expected residual correlations through its
proportional hazard approach. To preserve strict data pairing, whenever a data
set was disregarded as having produced a Heywood case, all five parallel data
set within the sample size were excluded. Complete datasets over all five FIV
loading levels were thus 930, 964 and 1000 for increasing sample sizes.

The initial model consisted in the logarithms of the design variables (expected
residual correlation and sample size). The complete model added FIV loading
(categorical) and its interactions. The fit improvement gave
$\chi^2(11)=55216$. This decomposes into the main effect of FIV loadings
($\chi^2(3)=41081$), and its interactions ($\chi^2(8)=14135$), all three tests
significant beyond 1e-15. All parameters indicated increased detection with
increasing FIV loadings, with increasing effect size and with increasing sample size. Figure~\ref{fig_2} depicts all detection rates along with the logistic curves 
adjusted by the GLM quasibinomial procedure. The detection rates are depicted by symbols corresponding to the FIV loading condition. Multiple points occur at equal effect sizes.

### Discussion
The present simulated data covered a wide range of four-variable conditions
(two FIVs and two dependent variables) crossed with five population FIV
correlations ranging from .20 to .72 and with 66 residual correlations from 0
to .34. Besides the expected beneficial effect of increased sample size, all
three characteristics inspected clearly stress the importance of using FIVs
that correlate highly when a choice is available. The nominal false positive
rate was respected in all combinations of test parameters but that with FIV
loadings of (.4, .5) along with sample size of 125. Similarly, variability of
residual correlation estimates about the corresponding population values
decreases markedly with increasing FIV loading while correct detection also
improved markedly. Finally, the residual correlation estimates were confirmed
to approach the corresponding population values as FIV loadings and sample size
increased, with no notable differences across the range of residual correlation
effect sizes.

# Application to Real Data

## Data

The above simulations were carried out with the FIVs already designated.
Important application details are worth adding through describing the residual
correlation analysis of a real data set. This will be done with data from
Ullman's (1996) chapter on Structural Equation Modelling (SEM) introduced in
the third edition of Tabachnick & Fidell textbook. She uses selected Wechsler
data from 175 learning disabled children. Eleven variables were extracted from
the WISC, the first six of which are initially taken as exclusive indicators of
Verbal IQ and the last five of Performance IQ. These are *a*: Information, *b*:
Comprehension, *c*: Arithmetic, *d*: Similarities, *e*: Vocabulary, *f*: Digit
span, *g*: Picture completion, *h*: Picture arrangement, *i*: Block design,
*j*: Object arrangement, and *k*: Coding. This initial model not being consistent
with the data ($\chi^2(43)=70.2, p=.0054$), the suggestion from LISREL-7
(Jöreskog & Sörbom, 1988) was accepted to add a cross-link from the Performance
factor to Comprehension, which yielded $\chi^2(42)=60.3, p=.033$. The further
removal of Coding, whose link from Performance was not significant, gave the
final model with $\chi^2(33)=45.0, p=.08$.

## FIV Selection

The R code to apply residual correlation analysis is available on GitHub. The
procedure consists in the following steps: (1) preprocess the data, (2) select
suitable FIVs, and (3) call the analysis.

Application of residual correlation analysis starts by subjecting the data to a
function called 'init_SCA'. The data could be a rectangular matrix with cases as
rows and variables as columns, or alternately a covariance or correlation
matrix along with sample size. For this example, the covariance matrix was
obtained from 'tf <- TFcovar(.R))' (see Listing \ref{lst:roundtbl}). The init_SCA function operates on the
correlation matrix, attempting signal cancellation on each pair of variables.
It returns a list, here named TF, with many elements, including the sequence of
tested pairs, the significance of their correlations, their optimised
cancellation weights and the associates $R^2$ probabilities. An ad hoc function
'roundTbl' is convenient to examine these results to help choosing suitable
FIVs. It prints a square matrix with cancellation $R^2$ probabilities above the
diagonal and, below, the weights of the optimised
contrast (applied to the highest ranked
variable before subtracting the lower ranked one from it, e.g. *-B+wX*). From this output, one can note that
the weights in the last row (the Coding scale) are markedly larger than all the
ones above. This is the signature of an orphan variable that is barely informed
by common factors. A perfectly independent variable already does not correlate with any remaining one. Signal cancellation optimisation then just accommodates random correlations.
Consistent with Coding having no correlation significant
at $p<.01$, this variable is excluded from further consideration.
 
Selection of FIVs for each factor starts by inspecting the table annotated by
roundTbl which adds a * after significant probabilities (default: seuil=.05) or
a = after probabilities above the threshold value expressed in negative (here
seuil=-.2). The latter flags the confident signal cancellations, i.e. contrasts
which multiple regression on all remaining variables could not explain, with a
security margin. FIV selection also considers the correlation matrix, given the
documented value of highly correlated FIVs.

Given the initial measurement model, the pair of exclusive Verbal factor
indicators should be selected among the first six variables. Comprehension and
Arithmetic (*b* and *c*) as well as Arithmetic and Similarities (*c* and *d*)
could not cancel their signal ($p=.009$ and $p=.033$). This means that at least
one of them is also informed by another common factor, which is not necessarily
Arithmetic. That Comprehension (*b*) cancelled signal with Similarities (*d*)
($p=.412$) requires that these two scales are well aligned with the signal
space origin. While exclusivity to a common factor achieves this alignment, so
can proportional loadings on a pair of factors. Thus, if Comprehension were to
be declared bifactorial, so should Similarities, a conclusion reinforced by
both these variables poorly cancelling the signal of the remaining four Verbal
intelligence variables. The doubt on the unifactorial nature of these three
variables leaves the remaining three Verbal variables to choose from as
indicators of this factor. Among these, the highest correlation observed is
.625 between Information and Vocabulary. These are retained as FIVs for the
Verbal IQ factor.

The four indicator variables of Performance allow six pairwise signal
cancellation among them. These show no counterindication (i.e. no significant
$R^2$) to their presumed FIV status, although four $R^2$ probabilities are
barely above the arbitrarily defined bar of .20. The remaining two are .717 and
.266 for signal cancellation of Block design (*i*) with Picture completion
(*g*) and with Object arrangement (*j*). This designates Block design as one
FIV. A higher non-significant probability (e.g., .717 vs .266), provided both are deemed reliable, does
not mean better signal cancellation. Since the simulations emphasised the
importance of high FIV loadings, preference for the second FIV is given to
Object arrangement that correlates with Block design .399 compared to .363 for
Picture completion.

## Results

Application of partitionne_R to the list produced by init_SCA is followed by
printing the main results with roundTbl using a threshold corrected for multiple tests (discussed below). Of the four initial
Verbal indicator other than the FIVs, *c* and *f* (Arithmetic and Digit span) have
no detectable signal outside the FIV signal direction. The other two variables,
*b* and *d* (Comprehension and Similarities) have no signal on the excluded
Verbal factor direction, but each has significant residual correlations on
three of the four indicators of the Performance factor (Coding being ignored).
With the Verbal factor direction excluded, all residual correlations among the
Performance indicators remain significant. Exclusion of the latter factor made
its remaining two variables behave as noise only while all Verbal indicators
still correlated significantly.
 
## Discussion

Partitioning the correlations among the ten retained variables gave two clear
indications that Comprehension and Similarities (*b* and *d*) are bifactorial.
First, the pattern of pairwise signal cancelation showed clear signal
cancellation success between these two variables but significant or nearly so
$R^2$ probabilities for signal cancellation of each with the remaining four
Verbal indicators. Second, the residual correlations showed that while the
removal of the Verbal factor direction from these two variables cancelled their
signal from the Verbal factor, they retained residual correlations with indicators
of the Performance factor.

While the textbook final solution acceptably fit the data when only
Comprehension was informed by both factors, with $\chi^2(33)=45.0, p=.08$, the
present factor exclusion strongly suggests that both Comprehension and
Similarities are bifactorial. Submitting this to lavaan (Rosseel, 2012) gave
$\chi^2(32)=37.7, p=.23$. The fit difference due to further acknowledging
Similarities as bifactorial is $\chi^2(1)=7.3, p=.007$. It is likely that the
extra link was also suggested by LISREL to further improve the already
statistically acceptable model, but parsimony would have argued against
incorporating it in the model. Had partitionne_R been available but not used
ahead of SEM, it could have provided useful information to decide if the
suggested extra link was necessary or simply opportunistic.

Having excluded Coding as unreliable, 28 residual correlations are tested for
each factor exclusion. Correcting for that many tests would, however, be
inappropriate since excluding the direction of one factor cannot be expected to
cancel shared information from the remaining factors. Residual correlations
among the variables exclusive to the other factor may only be reduced from
their original full signal space values due to factor correlations. 

Here, each
variable initially suspected of being exclusive to a given factor is of
separated interest. Even if it is bifactorial, its residual correlations with
the other variables of the same factor are expected null unless any of them is
also bifactorial. Three choices are relevant. Controlling for all residual
correlations of a tested variable is simple but conservative in the presence of other
variables exclusive to the excluded factor, which behave as noise. Excluding
these factor-exclusive variables, here *c* and *f* but not *b* and *d*, from the correction would be delicate and
possibly circular. The currently recommended correction is for the number of
residual correlations with variables associated with other factors. Using the
Bonferroni correction, for instance, would divide the nominal $\alpha$ level by
four for each Verbal variable outside this factor direction, and likewise by 6 for exclusion of the Performance factor direction.
Exclusion of the latter factor already had all six uncorrected probabilities
above .1, but a correction should apply for each variable of the Verbal factor.
The residual correlations of each with the four variables of the Performance
factor should be tested at .0125. 

Arithmetic (*c*) and Digit span (*f*) had
their smallest uncorrected probability above .09. That of Comprehension (*b*)
is .002 with the next one at .006. Similarities (*d*) is barely significant at .011. 
Disregarding the small values of two of the remaining three 
residual correlations probabilities, at .014 and .072, someone could argue that the test of its residual correlation with Comprehension should also have been counted, bringing the statistical threshold to .01,
and prefer risking missing the extra link. This requires discussing what would be the meaning of only this test being significant while all four residual correlations with variables from the other factor being clearly not significant. This would be the signature of a supplementary doublet factor expressed exclusively in *b* and *d*, hardly a parsimonious solution. This would further require discounting all significant zero-order correlations of Similarities with the Performance variables as exclusively due to correlations among factors. To exclude that a variable expresses both factors in the model, correcting for the number of variables of alternate factor(s) is recommended. The present interpretation is that both variables are very likely bifactorial.

# General Discussion

Pairwise signal cancellation allows testing the hypothesis that the signal parts
of two variables lie in the direction of the same factor. When this
same-direction property holds, their signal lengths may be equated and then
subtracted to cancel this signal direction forming a contrast consisting
exclusively of their combined noise which is expected orthogonal to any other
variable. The suitable contrast weight is obtained by minimising the multiple
regression of the contrast on all remaining variables. For safety, it is
suggested to accept successful signal cancellation only if the minimised $R^2$
has a probability above .20, although probabilities below .05 may be taken as
evidence that the two variables have their respective signal in a plane rather
than on a common direction.

Factor indicator variables, or FIVs, are pairs of variables initially suspected
exclusive to the same common factor and confirmed so by their successful signal
cancellation. A simple algebraic formula allows assessing the projection on
that factor of any other variable from its correlations with the two FIVs. The
hypothesis that two variables only share information from this common factor
implies that the product of their projections on the designated factor equates
their observed correlation except for accidental noise effects.

The difference between a correlation and the corresponding projection product
is designated a residual correlation. A suitable test of this difference
assesses the presence of shared common factor information outside the direction of the target
factor. While classical partial correlation assesses the correlation of two variables
within the variable space by excluding from each an optimal combination of
noisy variables, residual correlation selectively excludes the direction of a
factor suitably separated from the noise sources affecting the involved
variables. Partial correlation could not serve the present purpose of
validating that a pair of variables are exclusively informed by the same common
factor without error-free measurement of an independent variable
exclusive to the target factor, the violation of which is known to yield
inflated type I error rates (Brewer, Campbell & Crano, 1970; Brunner & Austin,
2009; Shear & Zumbo, 2013). Indeed, unreported partial correlations applied to
the simulated data, using the FIVs as independent variables, showed nearly 100%
false positives at N=2000. This not only reflects underestimation of the signal
contents of the correlated variables, but also injection into each of noise
from the independent variables.

The simulations documented the reliability of the test of residual correlations
in respecting the nominal false positive rate, except for some small excess
only with the smallest population FIV correlation of .20 along with the smallest sample size of
125, for which 633 extra data sets were required to produce 1000 for which the FIVs correlated significantly at the .05 level. They further documented that the assessed residual correlations
approximate well their expected values, with reducing variability as the
expected FIV correlation or the sample size increase.

Since a significant residual correlation negates that two variables are
exlcusive to the same factor, it may be used ahead of SEM to confirm which
variables are exclusive to a factor. This was exemplified using a textbook
dataset that illustrated SEM with fit improvement. The residual correlation
approach contradicted its final accepted solution by documenting that not just
one but two of the variables initially hypothesised exclusive to Verbal
intelligence are further informed by the Performance IQ factor.

The common-factor model conceptualises $n_f$ common factors informing $n_v$
variables beyond their $n_v$ source of unique variance. Because $n_f+n_v$
independent directions are expressed in only $n_v$ observables, any combination
of noisy variables meant as a factor remains within the variable space and
retains some noise. This does not prevent, say, exploratory factor analysis to
be useful but SEM further aims at recovering the missing $n_f$ dimensions,
separate from the $n_v$ unicities. Residual correlation similarly recovers
missing factor dimensions uncontaminated by combined sources of unique
variance, one factor at the time This is however limited to dimensions
expressed in at least two exclusive variables attested so by signal
cancellation.

Because factor exclusion projects all variables on the factor direction, the
effect of correlated factors is to reduce shared variance outside the excluded
direction. This was exemplified in the simulated data, where the pattern matrix
of correlated factor was expressed by a Cholesky decomposition placing the
target factor as the first axis with the residual loadings on correlated
factors on subsequent orthogonal axes.

A final point concerns the apparent relationship of tetrad differences
(Spearman & Holzinger, 1924, 1925) to the current assessment of projections on
a target factor. Both approaches combine observed correlations, whose expected
values are the sums of loading products over the orthogonal space of all common
factors. They also have in common to corroborate the single common factor
hypothesis by a non-significant statistic. But while the tetrad differences are
built to have null expected values over all subsets of four variables to
document that all correlations are due to a single common factor, this single
factor assumption is here only required for a pair of variables, the FIVs,
exclusive to each common factor of interest. This FIV assumption is directly
challenged by signal cancellation, which requires that the multiple regression
$R^2$ of their optimised contrast with the remaining variables is acceptably
null, for which a probability threshold of .2 is suggested. The similarity is
not as close as may seem at first sight.

## Significance

The proposed application of factor exclusion is to help decide which variables
are unifactorial. The main significance of the work reported here, however, is
the introduction of signal cancellation to isolate individual factors from the
unique sources of variance. Although SEM may already achieve this based on user
knowledge of the domain, the signal cancellation approach holds promise for
blind recovery of sparse factor structures, i.e. to recuperate without user
guidance each required factor loading and each required factor correlation
provided that each factor has at least two exclusive indicator variables
(Achim, 2024). The signal cancellation criterion used in this 2024 report
 relied on an unwarranted assumption, which following work corrected. Ongoing
work seeks to apply signal cancellation to also explain the signal of variables
contributed by factors without the required two FIVs. The direction of such
extra factors lacks adequate constraint for unambiguous relationships with the other factors but may be represented by the signal
direction of a representative variable.

# Data Availability Statement

The functions used, including code for the two figures, are available in the FactorExclusion package
(https://github.com/AndreAchim/FactorExclusion; DOI: 10.5281/zenodo.22319448),
which should eventually be incorporated into the SignalCancellation package
maintained by P.-O. Caron (Université TÉLUQ) on GitHub
(https://github.com/quantmeth/SignalCancellation). When completed, this package should be ported to CRAN. 

The simulation data may be reproduced in R by "out <- FE_tests()".
Function 'TFcovar' returns the WISC covariance matrix from Tabachnick & Fidell's textbook along with sample size.



# Acknowledgements

The author declares no conflict of interest but wishes to express gratitude for
a scholarship from Epilepsy-Canada/Parke-Davis, back in 1988, that allowed
pursuing his career in the development of innovative quantitative methods.

Anthropic Claude AI assistant provided support in suggesting, explaining and
implementing the statistical test for residual correlations and for the
conditional logistic regression from the Survival Analysis R package (Therneau,
2026). It also helped in preparing the figures and in formatting parts of this
manuscript besides providing valuable comments on a preliminary version.

\clearpage

# References

Achim, A. (2024). 'Blind recovery of sparse factor structures by signal cancellation', arXiv:2404.03781v2 [stat.ME].
https://doi.org/10.48550/arXiv.2404.03781.

Brewer, M. B., Campbell, D. T., & Crano, W. D. (1970). Testing a single-factor model as an alternative to the misuse of partial correlations in hypothesis-testing research. *Sociometry*, *33*(1), 1-11.
https://doi.org/10.2307/2786135

Brunner, L. J., & Austin, P. C. (2009). Inflation of type I error rate in
multiple regression when independent variables are measured with error.
*Canadian Journal of Statistics*, *37*(1), 33-46. doi:10.1002/cjs.10004

Jöreskog, K. G., & Sörbom, D. (1988). *LISREL 7. A Guide to the Program and Its Application*. Chicago: SPSS Inc.

Rosseel, Y. (2012). lavaan: An R package for structural Equation Modeling.
*Journal of Statistical Software*, *48*(2), 1–36.
https://doi.org/10.18637/jss.v048.i02

Shear, B. R., & Zumbo, B. D. (2013). False positives in multiple regression:
Unanticipated consequences of measurement error in the predictor variables.
*Educational and Psychological Measurement*, *73*(5), 733–756.
https://doi.org/10.1177/0013164412473358

Steiger, J. H. (1980). Tests for comparing elements of a correlation matrix.
*Psychological Bulletin*, *87*(2), 245–251.
https://doi.org/10.1037/0033-2909.87.2.245

Spearman, C. & Holzinger, K. (1924). The sampling error in the theory of two
factors. *The British Journal of Psychology*, *15*, 17-19.

Spearman, C. & Holzinger, K. (1925). Note on the Sampling Error of Tetrad
Differences. *The British Journal of Psychology*, *16*, 86-88.

Therneau, T. M. (2026). *A Package for Survival Analysis in R*.
https://CRAN.R-project.org/package=survival

Therneau, T. M., & Grambsch, P. M. (2000). *Modeling Survival Data: Extending the Cox Model*. Springer. https://doi.org/10.1007/978-1-4757-3294-8

Ullman, J.B. (1996). Structural Equation Modeling. In B. G. Tabachnick, & L. S.
Fidell (Eds.), *Using Multivariate Statistics* (3rd ed., pp. 709-819). New York,
NY: Harper Collins College Publishers.


\clearpage

```{r table1, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(knitr)
library(kableExtra)

tab1 <- data.frame(
  Variable = c("$v_1$", "$v_2$", "$v_3$", "$v_4$", "$v_5$", "$v_6$",
               "$v_7$", "$v_8$", "$v_9$", "$v_{10}$", "$v_{11}$",
               "$v_{12}$", "$v_{13}$", "$v_{14}$"),
  A1 = c(".4 to .8", ".5 to .9", ".40", ".74", ".40", "-.40",
         ".50", ".50", ".80", ".60", ".46", ".00", ".54", ".73"),
  A2 = c(".00", ".00", ".00", ".27", ".55", ".50", ".00", ".00",
         ".30", ".35", ".60", ".30", ".40", ".30"),
  A3 = c(".00", ".00", ".00", ".00", ".00", ".00", ".60", ".40",
         ".40", ".30", ".40", ".40", "-.30", ".40")
)

kable(tab1,
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      centering = FALSE,
      align     = c("l", "r", "r", "r"),
      col.names = c("Variable", "I", "II", "III"),
      caption   = "Variable coordinates of the test factor structure expressed on three orthogonal axes, the first of which, identified by variables \\protect$v_1$ and \\protect$v_2$, is to be excluded from the remaining variables.",
      label     = "structure") |>
  add_header_above(c(" " = 1, "Orthogonal axes" = 3))
```

\clearpage

\begin{figure}[H]
% \includegraphics[width=1\linewidth, height=0.88\textheight, keepaspectratio]{Figure1.pdf}
\caption{Observed means ± 1 standard deviation of the 46 test residual correlations superimposed on their expected values for three sample sizes and five loading levels of the two factor indicator variables.}
\label{fig_1}

\medskip
\noindent\textit{Alt text: The top-left of 15 panels is for the FIV loadings of .4, .5, with N=125. The white line representing estimated residual correlations is notably jittered around the black line of expected values but nearly superimposes it at higher sample sizes, even for this smallest FIV correlation. Some jitter remains for le .5, .6 loadings at N=125. The vertical width of the grey ribbon depicting plus-minus one standard deviation of estimates decreases markedly across the panels of increasing sample size and slightly across increasing FIV loadings.}
\end{figure}

\begin{figure}[H]
% \includegraphics[width=1\linewidth, height=0.88\textheight, keepaspectratio]{Figure2.pdf}
\caption{Detection rates and best fitted curves across increasing population residual correlation values by loading level of the pair of factor indicator variables for each sample size.}
\label{fig_2}

\medskip
\noindent\textit{Alt text: TPanels for each sample size show the fitted detection success across increasing effect size for the five FIV loading level. At N=125, only the higher FIV loadings at the largest effect size approaches 100\% success. At N=500, the curves asymptote 100\% near effect sizes of about .16 for the largest loading, up to the upper limit of .34 for the .5, .6 set of loadings, while the lowest FIV loading level nears 95\% at the largest effect size. At N=2000, this condition reaches ceiling near effect size of about .25, where the largest three sets of loadings do so by .12 and the fourth by .18.}
\end{figure}

\clearpage

```{=latex}
\begin{lstlisting}[language={}, basicstyle=\scriptsize\ttfamily, caption={Commands and outputs in applying factor exclusion to the illustrative data.}, label=lst:roundtbl]
> tf <- TFcovar()

> TF <- init_SCA(tf$TFdat,tf$N)

> roundTbl(TF$mPaires,3,seuil=-.2)
    a	    b       c     d      e      f      g      h     I        j      k
a   ~~~    0.072  0.135  0.235= 0.166  0.531= 0      0.001  0       0      0.271=
b  1.088    ~~~   0.009  0.412= 0.117  0.096  0.075  0.164  0.008   0.003  0.102
c  1.404   1.26    ~~~   0.033  0.535= 0.776= 0.001  0.008  0       0.024  0.125
d  1.083   0.997_ 0.815   ~~~   0.052  0.173  0.022  0.01   0.021   0.001  0.172
e  0.998   0.912  0.786_ 0.908   ~~~   0.501= 0      0.028  0       0      0.163
f  2.016   1.966  1.448_ 1.904  1.897_  ~~~   0.002  0.008  0.016   0.002  0.34=
g  1.634   1.355  1.12   1.37   1.63   0.491   ~~~   0.198  0.221=  0.717= 0.22=
h  2.483   1.681  1.619  2.228  2.078  0.79   1.29    ~~~   0.235=  0.201= 0.097
i  1.504   1.248  0.999  1.231  1.715  0.534  0.918_ 0.652_  ~~~    0.266= 0.144
j  2.883   1.609  1.23   1.679  1.845  0.542  1.073_ 0.831_ 1.211_   ~~~   0.099
k  5.746_ 10.196  7.892 39.756 26.308 10.467_ 6.503_ 6.448  7.816  39.585   ~~~

> tf15 <- partitionne_R(TF,c(1,5))

> roundTbl(tf15$corr_resid,3,seuil=.05/4)
    b       c      d      f      g      h       i      j
b   ~~~    0.462  0.111  0.764  0.002* 0.378   0.012* 0.006*
c  0.044    ~~~	  0.98   0.479  0.709  0.097   0.172  0.182
d  0.091   0.002   ~~~   0.887  0.014  0.011*  0.498  0.072
f -0.018   0.046 -0.009   ~~~   0.415  0.36    0.373  0.39
g  0.203_ -0.024  0.154 -0.055   ~~~   0.012*  0*     0*
h  0.056   0.112  0.161_ 0.065  0.182_  ~~~    0*     0.005*
i  0.161_  0.089  0.042 -0.06   0.275_ 0.283_   ~~~   0*
j  0.175_ -0.087  0.113 -0.06   0.288_ 0.205_  0.322_  ~~~

> tf910 <- partitionne_R(TF,c(9,10))

> roundTbl(tf910$corr_resid,3,seuil=.05/6)
    a       b      c      d      e      f       g      h
a   ~~~    0*     0*     0*     0*     0*      0.612  0.525
b  0.289_   ~~~   0.001* 0*     0*     0.011*  0.273  0.336
c  0.439_  0.299_  ~~~   0*     0*     0.001*  0.561  0.104
d  0.377_  0.281_ 0.298_  ~~~   0*     0.003*  0.11   0.187
e  0.504_  0.328_ 0.324_ 0.382_  ~~~   0*      0.382  0.561
f  0.32_   0.192_ 0.255_ 0.226_ 0.265_  ~~~    0.708  0.148
g  0.038   0.085  0.054  0.122  0.066  0.028    ~~~   0.693
h  0.048  -0.071  0.146  0.1   -0.043  0.111  -0.03    ~~~
\end{lstlisting}
```
